The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Nature or tidal nurture?

### 2025-10-08 thesis/extra_results.tex
grammar/spelling edits

\chapter{Additional simulation tests}\label{sec:extra_results}

In this section, we briefly explore additional simulations testing
assumptions not varied in the main text: variations in the halo
concentration, cored halos, alternative orbits, anisotropy, and
ellipsoidal models. While each of these variables influences total dark
matter evolution, the resulting inner structure is similar across
models. We conclude that consideration of these effects likely would not
change our conclusions.

@@

The following models aim to reproduce the stellar velocity dispersion of
Scl to within \(\lesssim 1\,\kms\) with a similar present-day half-light
radius.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_halo_boundmass.pdf}}
\caption[Tidal dependence on halo concentration]{A comparison of the
evolution of different N-body models for different halo concentrations
(heavier and lighter halo), and the heavy halo with the action-angle
corrected orbit. In the left, the maximum circular velocity \(\vmax\) of
the dark matter halo is plotted as a function of simulation time. In the
right, \(\vmax\) is instead plotted as a function of
\(\rmax\).}\label{fig:tidal_tracks_concentration}
\end{figure}

@@

In Fig.~\ref{fig:tidal_tracks_structure}, we show the evolution of the
cored model as compared to a heavier halo. The cored halo loses more
total dark matter mass, evolving further along the tidal track as
expected. However, in order to reproduce the observed velocity
dispersion, the final halo maintains a higher \(\vmax\) than the
fiducial halo.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/cored_density_i_f.pdf}}
\caption[Cored initial and final density profile]{The initial (dotted
orange) and final (solid orange) 3D density profiles for the cored model
of Scl on the \smallperi{} orbit. The blue thin line represents an NFW
halo with the same \(M_s\) and \(r_s\).}\label{fig:cored_i_f}
\end{figure}

\subsection{Velocity Anisotropy}\label{velocity-anisotropy}

Radial velocity anisotropy may cause halos to disrupt faster
\citep[e.g.][]{chiang+bosch+schive2024}. To test the effects of moderate
velocity anisotropy, we initialize a model with a velocity anistropy
with an Osipkov-Merritt profile rising from \(\beta=0.2\) at the centre
to \(\beta=1\) and infinity, with scale length \(4r_s\).

Fig.~\ref{fig:anisotropy_i_f} shows the initial and final anisotropy
profiles. The dwarf galaxy becomes more isotropic with tidal evolution.
Particles on more radially-anisotropic orbits are more easily stripped
as they have larger apocentres than more circular orbits of the
equivalent energy. Regardless, the overall tidal evolution is very
similar to our fiducial, isotropic case (see
Fig.~\ref{fig:tidal_tracks_structure}).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
\caption[Tidal evolution of anisotropy]{The initial and final anisotropy
profiles for the initially-anisotropic model of Scl on the \smallperi{}
orbit. The final profile is far more isotropic (\(\beta=0\)) than the
initial.}\label{fig:anisotropy_i_f}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_structure_boundmass.pdf}}
\caption[Tidal dependence on halo structure]{Similar to
Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
including a core, velocity anisotropy, and evolving an oblate halo.
While the normalization may differ, the tidal evolution is
similar.}\label{fig:tidal_tracks_structure}
\end{figure}

@@

Sculptor and Ursa Minor may be highly elliptical in 3D, possibly
violating the assumption of spherical symmetry
\citep[e.g.,][]{an+koposov2022}. The shape of the underlying dark matter
halo is unknown, however.

We create an oblate initial dark matter halo using \agama{}.
Fig.~\ref{fig:oblate_i_f} shows the initial equilibrium (after 5Gyr in
isolation) isodensity contours of our model. The initial snapshot is
sampled from the distribution function \begin{equation}{
\begin{split}
f = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h(J)}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g(J)}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g(J)}{J_{\rm cutoff}} \right)^\zeta\right] 
\end{split},
}\end{equation} (\texttt{doublePowerLaw} in \agama{}), with
\(g(J) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
\(h(J) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
\texttt{example\_doublepowerlaw.py} script in \agama{} solves for the
best parameters matching a given density profile. We chose to create a
model resembling an NFW but scaled by a factor 0.5 in the \(z\) axis.
For a scale-free halo (\(r_s=1\), \(M_s=1\)), the best-fit parameters
are \(J_0=0.890\), \(\Gamma=1.46\), \(\eta=0.568\), \(B=2.97\),
\(h_r=0.845\), \(h_z=1.66\), \(g_r=0.753\), \(g_z=1.69\), \(M_0=0.965\),
\(J_{\rm cutoff}=2.40\), \(\zeta=20\). The resulting initial conditions
are stable in isolation and produce a halo with near the desired density
profile and ellipticity. We scale the halo to have a major axis density
profile equivalent to an NFW with \(\rmax = 7\,\kpc\) and
\(\vmax=48\,\kpc\).

We find the evolution of the oblate halo to be nearly identical to the
spherical halo (see Fig.~\ref{fig:tidal_tracks_structure}), similar to
\citet{battaglia+sollima+nipoti2015}. The oblate halo however becomes
spherical after tidal evolution (Fig.~\ref{fig:oblate_i_f}). If Scl's
dark matter halo is indeed elliptical today, this model may not be an
adequate description.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_projected_2d.pdf}}
\caption[Oblate halo before and after tidal evolution]{The initial and
final (after 9Gyr of tidal evolution) projected density profiles for the
oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\) and
\(z'\) the major and minor axes. The contours are drawn assuming normal
smoothing of 0.4 kpc and are log-spaced with intervals of 0.105
dex.}\label{fig:oblate_i_f}
\end{figure}

@@

Because of the strong tidal interaction with the MW, the trajectory is
substantially perturbed from a point orbit. We adjust the initial
conditions by comparing the change in actions (as calculated in the
static MW-only potential) before and after the pericentre. Our final
model, the \texttt{MW\ impact} model, is able to approximately reproduce
the observed position of Scl.

Fig.~\ref{fig:tidal_tracks_umi} compares the tidal evolution of Scl on
the mean, \texttt{MW\ impact} , and \smallperi{} orbit. The mean orbit
evloves less down the tidal track than the \smallperi{} model. Instead,
the \texttt{MW\ impact} orbit experiences most tidal evolution during
its first MW-pericentre. While evolving further along the tidal track,
the stars of this model nevertheless remain exponential
(Fig.~\ref{fig:scl_mw_impact_i_f}). We suggest that the impulsive
pericentric passage does not occur for long enough in this model to
produce the expected extended density profile. A yet more extreme
orbital history would be necessary to tidally transform Scl's stars.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_orbits_boundmass.pdf}}
\caption[Ursa Minor tidal dependence on orbit]{Similar to
Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
orbits in the LMC. Most LMC models evolve more weakly than the MW
models, however these do not reach full agreement with the present-day
position so misrepresent recent tidal
evolution.}\label{fig:tidal_tracks_umi}
\end{figure}

@@

Fig.~\ref{fig:scl_tidal_stream} and Fig.~\ref{fig:umi_tidal_stream}
shows the resulting distributions of velocities and distance along the
stream orbital axis \(\xi'\) for Scl and UMi. The tidal tails may have
detectible gradients in radial velocities (\(\sim 10\,\kms\)) and proper
motions (mostly for UMi, of \(\sim 0.1\,\masyr\)). However, detecting
such a gradient would require tracing stars across several degrees on
the sky.

>>>>

\chapter{Additional Simulation Results}\label{sec:extra_results}

In this section, we briefly explore additional simulations testing
variations in the halo concentration, cored halos, alternative orbits,
anisotropies, and ellipticities. While each of these variables
influences total dark matter evolution, the resulting inner structure is
similar across models. We conclude that consideration of these effects
likely would not change our conclusions.

@@

The following models aim to reproduce the stellar velocity dispersion of
Sculptor to within \(\lesssim 1\,\kms\) with a similar present-day
half-light radius.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_halo_boundmass.pdf}}
\caption[Tidal dependence on halo concentration]{A comparison of the
evolution of different N-body models for different halo concentrations
(heavier and lighter halo), and the heavy halo with the action-angle
corrected orbit. \textbf{Left:} The maximum circular velocity \(\vmax\)
of the dark matter halo is plotted as a function of simulation time.
\textbf{Right:} \(\vmax\) is instead plotted as a function of
\(\rmax\).}\label{fig:tidal_tracks_concentration}
\end{figure}

@@

In Fig.~\ref{fig:tidal_tracks_structure}, we show the evolution of the
cored model as compared to a heavier halo. The cored halo appears to
evolve more mildly than the NFW halo. This is possibly because, to match
the velocity dispersion, the cored halo has \(\sim 50\%\) more mass
within \(1\,\kpc\) than a similar cuspy halo.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/cored_density_i_f.pdf}}
\caption[Tidal evolution of a cored density profile]{The initial (dotted
orange) and final (solid orange) 3D density profiles for the cored model
of Scl on the \smallperi{} orbit. The blue thin line represents an NFW
halo with the same \(M_s\) and \(r_s\).}\label{fig:cored_i_f}
\end{figure}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_mw_structure_boundmass.pdf}}
\caption[Tidal tracks depending on halo substructure]{Similar to
Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
including a core, velocity anisotropy, and evolving an oblate halo.
While the normalization may differ, the tidal evolution is
similar.}\label{fig:tidal_tracks_structure}
\end{figure}

\subsection{Velocity anisotropy}\label{velocity-anisotropy}

Radial velocity anisotropy may cause halos to disrupt faster
\citep[e.g.][]{chiang+bosch+schive2024}. To test the effects of moderate
velocity anisotropy, we initialize a model with a velocity anisotropy
with an Osipkov-Merritt profile rising from \(\beta=0.2\) at the centre
to \(\beta=1\) and infinity, with scale length \(4r_s\).

Fig.~\ref{fig:anisotropy_i_f} shows the initial and final anisotropy
profiles after tidal evolution on Scl's \smallperi{} orbit. The dwarf
galaxy becomes more isotropic with tidal evolution. Particles on more
radially anisotropic orbits are more easily stripped as they have larger
apocentres than more circular orbits of the equivalent energy.
Regardless, the overall tidal evolution is very similar to our fiducial,
isotropic case (see Fig.~\ref{fig:tidal_tracks_structure}).

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/anisotropy_i_f.pdf}}
\caption[Tidal evolution of anisotropy]{The initial and final anisotropy
profiles for the initially anisotropic model of Scl on the \smallperi{}
orbit. The final profile is far more isotropic (\(\beta=0\)) than the
initial.}\label{fig:anisotropy_i_f}
\end{figure}

@@

Sculptor and Ursa Minor may be highly elliptical in 3D, possibly
violating the assumption of spherical symmetry
\citep[e.g.,][]{an+koposov2022}. Although, the shape of the underlying
dark matter halo is unknown.

We create an oblate initial dark matter halo using \agama{}.
Fig.~\ref{fig:oblate_i_f} shows the initial equilibrium (after 5Gyr in
isolation) isodensity contours of our model. The initial snapshot is
sampled from the distribution function \begin{equation}{
\begin{split}
f = \frac{M_0}{(2\pi\,J_0)^3} \left[1 + \left(\frac{J_0}{h(J)}\right)^\eta\right]^{\Gamma / \eta} \ \left[1 + \left(\frac{g(J)}{J_0}\right)^\eta\right]^{-B/\eta} \\ \ \exp\left[-\left(\frac{g(J)}{J_{\rm cutoff}} \right)^\zeta\right] 
\end{split},
}\end{equation} (\texttt{doublePowerLaw} in \agama{}), with
\(g(J) = g_r J_r + g_z J_z + (3-g_r - g_z) |J_\phi|\) and
\(h(J) = h_rJ_r + h_zJ_z + (3-h_r-h_z) |J_\phi|\). The
\texttt{example\_doublepowerlaw.py} script in \agama{} solves for the
best parameters matching a given density profile. We chose to create a
model resembling an NFW but scaled by a factor of 0.5 in the \(z\)-axis.
For a scale-free halo (\(r_s=1\), \(M_s=1\)), the best-fit parameters
are \(J_0=0.890\), \(\Gamma=1.46\), \(\eta=0.568\), \(B=2.97\),
\(h_r=0.845\), \(h_z=1.66\), \(g_r=0.753\), \(g_z=1.69\), \(M_0=0.965\),
\(J_{\rm cutoff}=2.40\), \(\zeta=20\). The resulting initial conditions
are stable in isolation and produce a halo with near the desired density
profile and ellipticity. We scale the halo to have a major axis density
profile equivalent to an NFW with \(\rmax = 7\,\kpc\) and
\(\vmax=48\,\kpc\).

We find the evolution of the oblate halo to be nearly identical to the
spherical halo (see Fig.~\ref{fig:tidal_tracks_structure}), similar to
\citet{battaglia+sollima+nipoti2015}. The oblate halo becomes spherical
after tidal evolution (Fig.~\ref{fig:oblate_i_f}). If Scl's dark matter
halo is indeed elliptical today, this model may not be an adequate
description.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/oblate_projected_2d.pdf}}
\caption[Oblate halo projected density snapshots]{The initial and final
(after 9Gyr of tidal evolution) projected density profiles for the
oblate halo on Scl's \smallperi{} orbit, projected on with \(x'\) and
\(z'\) the major and minor axes. The contours are drawn assuming normal
smoothing of 0.4 kpc and are log-spaced with intervals of 0.105
dex.}\label{fig:oblate_i_f}
\end{figure}

@@

Because of the strong tidal interaction with the MW, the trajectory is
substantially perturbed from a point orbit. We adjust the initial
conditions by comparing the change in actions (as calculated in the
static MW-only potential) before and after the pericentre. Our final
model, the \texttt{MW\ impact} model, can approximately reproduce the
observed position of Scl.

Fig.~\ref{fig:tidal_tracks_umi} compares the tidal evolution of Scl on
the mean, \texttt{MW\ impact}, and \smallperi{} orbit. The mean orbit
loses less mass than the \smallperi{} model. Instead, the
\texttt{MW\ impact} orbit experiences most tidal evolution during its
first MW pericentre. While evolving further along the tidal track, the
stars of this model nevertheless remain exponential
(Fig.~\ref{fig:scl_mw_impact_i_f}). We suggest that the impulsive
pericentric passage does not occur for long enough in this model to
produce the expected extended density profile. A yet more extreme
orbital history would be necessary to tidally transform Scl's stars.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_orbits_boundmass.pdf}}
\caption[Sculptor's tidal evolution for different orbits]{Similar to
Fig.~\ref{fig:tidal_tracks_concentration}, except testing the effects of
orbits in the LMC. Most LMC models evolve more weakly than the MW
models.}\label{fig:tidal_tracks_umi}
\end{figure}

@@

Fig.~\ref{fig:scl_tidal_stream} and Fig.~\ref{fig:umi_tidal_stream} show
the resulting distributions of velocities and distance along the stream
orbital axis \(\xi'\) for Scl and UMi. The tidal tails may have
detectible gradients in radial velocities (\(\sim 10\,\kms\)) and proper
motions (mostly for UMi, of \(\sim 0.1\,\masyr\)). However, detecting
such a gradient would require tracing stars across several degrees on
the sky.

### 2025-10-08 thesis/extra_rv_models.tex
grammar/spelling edits

\chapter{Line-of-sight velocities: Sample selection and
modelling}\label{sec:extra_rv_models}

In this section, we analyze the observed line-of-sight (LOS) velocity
distributions for Sculptor (Scl) and Ursa Minor (UMi). Our aim is to
test for kinematic tidal signatures by combining literature LOS
velocities with J+24's membership framework. Our derived systemic
velocities and dispersions are consistent with literature. We find weak
evidence for a velocity gradient in Scl. As the gradient is misaligned
with the orbit, the gradient more likely reflects intrinsic rotation
than tidal disruption. Scl's velocity dispersion also rises within the
effective radius, but predominantly in the inner regions, contrary to
tidal disruption. We find no evidence of a gradient in mean velocity or
velocity dispersion in UMi. We conclude that Scl and UMi do not show
clear features of tidal disruption given current velocity observations.

@@

We compile LOS velocities from several spectroscopic surveys. For
Sculptor, we combine \citet{tolstoy+2023}; \citet{walker+2009};
\citet{sestito+2023a}; and APOGEE \citep[DR17,][]{abdurrouf+2022}. For
Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
\citet{sestito+2023b}; and APOGEE. We then crossmatch all catalogues to
J+24 Gaia stars. If a study did not report GaiaDR3 source ID's, we match
to the nearest star within 1--3 arcseconds. We combine measurements of
the same star using inverse-variance weighting. To reduce likely
binaries, we remove stars with significant velocity
dispersions.\footnote{Specifically, using that
  \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
  \(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
  distribution with \(N-1\) measurements.}

We build on J+24's likelihood by adding multiplicative terms in the
total likelihood for the velocity consistency (see
Section~\ref{sec:the_algorithm}). We assume that the satellite and
background \(v_{\rm los, gsr}\) distributions are Gaussian. We assume a
halo/background velocity dispersion of a constant
\(\sigma_{\rm halo} = 100\,\kms\) with mean 0
\citep[e.g.][]{brown+2010}. We select stars with velocity-informed
satellite membership probabilities of greater than 0.2. For Scl, we find
1918 unique members and UMi, 831.

@@

We correct the velocities for the solar motion and on-sky size of the
galaxy. We transform the velocities into the galactic standard of rest
(GSR, i.e.~same location as ICRS but velocities relative to the galactic
centre), and then transform velocities to correct for the apparent
gradient induced by the dwarf's proper motion \citep[see
also][]{WMO2008, strigari2010}. We define \(v_{\rm gsr}'\) to be
velocities in the GSR frame subtracting the PM-induced gradient. The
correction from both effects induces an apparent gradient of about
\(1.3\,\kmsdeg\) for Sculptor and less for Ursa Minor. The uncertainty
on this velocity correction are less than the individual star
uncertainties and any derived velocity gradients.

\section{MCMC modeling}\label{mcmc-modeling}

We fit Monte-Carlo Markov chain (MCMC) models to solve for the systemic
velocity, velocity dispersion, and possible gradients in velocities or
velocity dispersions. We assume that the galaxy follows a planar
velocity gradient. The mean velocity at a given point on the sky,
\(\mu\), is assumed to be,

\begin{equation}{
\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta,
}\end{equation}

for tangent-plane coordinates \(\xi\) and \(\eta\), systemic velocity
\(\mu_0\), and velocity gradient slopes \(a\) and \(b\). The velocity
dispersion at a given position, \(\sigma\), is assumed to depend as a
power-law on elliptical radius \(R_{\rm ell}\) alone: \begin{equation}{
\log \sigma = \log \sigma_0 + c\,\log(R_{\rm ell} / R_h)
}\end{equation} where \(\sigma_0\) is the system's velocity dispersion
at \(R_h\), and \(c\) is the velocity dispersion gradient slope. We use
weakly-informative priors, as described in Table~\ref{tbl:scl_rv_mcmc},
and with \(a, b \sim N(0, 6^2)\,\kmsdeg\).

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
\caption[LOS velocity fit to Scl]{\textbf{Left}: Velocity histogram of
Scl and UMi in terms of projected-correction GSR LOS velocity. Black
points with error bars are from the crossmatched observed sample, and
green lines represent MCMC fits to the velocity dispersion.
\textbf{Right} The observed velocity dispersion gradient (black points)
in 10 equal number bins and the derived slopes for MCMC samples (green
lines). Scl shows moderate evidence for an increasing velocity
dispersion with radius, but Scl's velocity dispersion appears to flatten
outside of \(R_h\). UMi's evidence for a radial gradient is
weaker.}\label{fig:rv_hists}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter_gradient.png}}
\caption[Scl velocity gradient]{\textbf{Top} members of Sculptor plotted
in the tangent plane coloured by corrected velocity difference from mean
\(v_z - \bar v_z\) . The black ellipse marks the half-light radius in
Fig.~\ref{fig:scl_selection}. The black and green arrows mark the proper
motion (PM, GSR frame) and derived velocity gradient (rot) vectors (to
scale). \textbf{Bottom}: The corrected LOS velocity along the best fit
rotational axis. RV members are black points, the systematic \(v_z\) is
the horizontal grey line, blue lines represent the (projected) gradient
from MCMC samples, and the orange line is a rolling median (with a
window size of 50).}\label{fig:scl_velocity_gradient_scatter}
\end{figure}

We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\)with
velocity dispersion \(9.64\pm0.16\,\kms\) (see Fig.~\ref{fig:rv_hists}).
Our values are very consistent with previous work
\citep[e.g.][\citet{arroyo-polonio+2024},
\citet{battaglia+2008}]{walker+2009}. We detect a moderately significant
gradient of \(4.3\pm1.3\,\kmsdeg\) at a position angle of
\(-149_{-13}^{+17}\) degrees, similar to past detections
\citetext{\citealp[e.g.,][]{arroyo-polonio+2024}; \citealp{battaglia+2008}; \citealp[but
see also][]{strigari2010}; \citealp{martinez-garcia+2023}}.

Fig.~\ref{fig:scl_velocity_gradient_scatter} plots the combined velocity
sample and the MCMC samples for the velocity gradient. While some
samples have a consistent gradient with 0, most samples have a positive
velocity gradient, consistent with the rolling median trend. The
velocity gradient in Scl is misaligned with the proper motion

@@

\emph{Binarity}. Stars in binary systems can inflate the inferred
system's velocity dispersion. For Scl and UMi with high measured
velocity dispersion, binaries likely add \(\sim 1\,\kms\) to the
velocity dispersion \citep{spencer+2018, gration+2025}.

\emph{Multiple populations}. Both Sculptor and Ursa Minor likely contain
multiple populations with different kinematics\citep[\citet{pace+2020},
\citet{tolstoy+2004}]{arroyo-polonio+2024}. We do not model these
separately---it is unclear how to define a velocity dispersion in such a
case.

@@

\emph{Tangental motions.} Precise star-by-star proper motions would
further test for kinematic disequilibrium. Presently, \emph{Gaia}
uncertainties are too large to permit such studies for Sculptor and Ursa
Minor.\footnote{Specifically, typical proper motion uncertntanties for
  faint, member stars of Scl and UMi are around \(0.5\,\masyr\),
  corresponding to velocities \(\sim 100\,\kms\).}

@@

In this section, we analyze literature samples of LOS velocity
measurements for both Scl and UMi. In each case, we find systemic LOS
velocities and velocity dispersions consistent with past work. We detect
weak evidence for a velocity gradient in Scl, however the gradient is
mis-aligned with the proper motion of Scl so is unlikely to be of tidal
origin. In both galaxies, the velocity dispersion profile is consistent
with a constant velocity dispersion with radius. The lack of
observational evidence for ongoing tidal disruption in the velocity
distribution of stars in Scl and UMi further supports our
interpretations in the main text.

>>>>

\chapter{Line-of-Sight Velocities: Sample Selection and
Modelling}\label{sec:extra_rv_models}

In this section, we analyze the observed line-of-sight (LOS) velocity
distributions for Sculptor (Scl) and Ursa Minor (UMi). We aim to test
for kinematic tidal signatures by combining literature LOS velocities
with J+24's membership framework. Our derived systemic velocities and
dispersions are consistent with the literature. We find weak evidence
for a velocity gradient in Scl. As the gradient is misaligned with the
orbit, the gradient more likely reflects intrinsic rotation than tidal
disruption. Scl's velocity dispersion also rises within the effective
radius, but predominantly in the inner regions, contrary to tidal
disruption. We find no evidence of a gradient in mean velocity or
velocity dispersion in UMi. We conclude that Scl and UMi do not show
clear features of tidal disruption given current velocity observations.

@@

We compile LOS velocities from several spectroscopic surveys. For
Sculptor, we combine \citet{tolstoy+2023}; \citet{walker+2009};
\citet{sestito+2023a}; and APOGEE \citep[DR17,][]{abdurrouf+2022}. For
Ursa Minor, we combine \citet{spencer+2018}; \citet{pace+2020};
\citet{sestito+2023b}; and APOGEE. We then cross-match all catalogues to
J+24 Gaia stars. If a study did not report \emph{Gaia} DR3 source IDs,
we match to the nearest star within 1--3 arcseconds. We combine
measurements of the same star using inverse-variance weighting. To
reduce likely binaries, we remove stars with significant velocity
dispersions.\footnote{Specifically, using that
  \(\chi^2=\frac{s^2}{\delta \bar v^2}\), we remove stars with a
  \(\chi^2\) larger than the 99.9th percentile of the \(\chi^2\)
  distribution with \(N-1\) measurements.}

We build on J+24's likelihood by adding multiplicative terms in the
total likelihood for the velocity consistency (see
Section~\ref{sec:the_algorithm}). We assume that the satellite and
background \(v_{\rm los, gsr}\) distributions are Gaussian in the
Galactic Standard of Rest (GSR, i.e., same location as ICRS but
velocities relative to the galactic centre). For the satellite, we adopt
a mean and standard deviation based on
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}, and, for the
background, mean \(0\,\kms\) and dispersion
\(\sigma_{\rm halo} = 100\,\kms\) \citep[e.g.][]{brown+2010}. We select
stars with velocity-informed satellite membership probabilities of
greater than 0.2. For Scl, we find 1918 unique members and UMi, 831.

@@

We correct the velocities for the solar motion and the on-sky size of
the galaxy. We transform the velocities into the GSR and correct for the
apparent gradient induced by the dwarf's proper motion
\citep[see,][]{WMO2008, strigari2010}. We define \(v_{\rm gsr}'\) to be
velocities in the GSR frame, subtracting the PM-induced gradient. The
correction from both effects induces an apparent gradient of about
\(1.3\,\kmsdeg\) for Sculptor and less for Ursa Minor. The uncertainty
on this velocity correction is less than the individual star
uncertainties and any derived velocity gradients.

\section{Monte Carlo Markov chain
modelling}\label{monte-carlo-markov-chain-modelling}

We fit Monte Carlo Markov chain (MCMC) models to solve for the systemic
velocity, velocity dispersion, and possible gradients in velocities or
velocity dispersions. We assume that the galaxy follows a planar
velocity gradient. The mean velocity at a given point on the sky,
\(\mu\), is assumed to be,

\begin{equation}{
\mu(\xi, \eta) = \mu_0 + a\,\xi + b\,\eta,
}\end{equation} for tangent plane coordinates \(\xi\) and \(\eta\),
systemic velocity \(\mu_0\), and velocity gradient slopes \(a\) and
\(b\). The velocity dispersion at a given position, \(\sigma\), is
assumed to depend as a power-law on elliptical radius \(R_{\rm ell}\)
alone: \begin{equation}{
\log \sigma = \log \sigma_0 + c\,\log(R_{\rm ell} / R_h)
}\end{equation} where \(\sigma_0\) is the system's velocity dispersion
at \(R_h\), and \(c\) is the velocity dispersion gradient slope. We use
weakly-informative priors, as described in Table~\ref{tbl:scl_rv_mcmc},
and with \(a, b \sim N(0, 6^2)\,\kmsdeg\).

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_umi_rv_fits.pdf}}
\caption[Velocity dispersion fits]{\textbf{Left:} Velocity histogram of
Scl and UMi in terms of projected-correction GSR LOS velocity. Black
points with error bars are from the sample of LOS-velocity members, and
green lines represent MCMC fits to the velocity dispersion.
\textbf{Right:} The observed velocity dispersion gradient (black points)
in 10 equal number bins and the derived slopes for MCMC samples (green
lines). Scl shows moderate evidence for an increasing velocity
dispersion with radius, but Scl's velocity dispersion appears to flatten
outside of \(R_h\). UMi's evidence for a radial gradient is
weaker.}\label{fig:rv_hists}
\end{figure}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/scl_rv_scatter_gradient.png}}
\caption[A possible velocity gradient in Sculptor]{\textbf{Top} members
of Sculptor plotted in the tangent plane, coloured by corrected velocity
difference from mean \(v_z - \bar v_z\). The black ellipse marks the
half-light radius in Fig.~\ref{fig:scl_selection}. The black and green
arrows mark the proper motion (PM, GSR frame) and derived velocity
gradient (rot) vectors (to scale). \textbf{Bottom}: The corrected LOS
velocity along the most likely rotational axis. RV members are black
points, the systematic \(v_z\) is the horizontal grey line, blue lines
represent the (projected) gradient from MCMC samples, and the orange
line is a rolling median (with a window size of
50).}\label{fig:scl_velocity_gradient_scatter}
\end{figure}

We derive a systemic velocity for Sculptor of \(111.3\pm0.2\,\kms\) with
velocity dispersion \(9.64\pm0.16\,\kms\) (see Fig.~\ref{fig:rv_hists}).
Our values are very consistent with previous work
\citep[e.g.,][]{walker+2009, arroyo-polonio+2024, battaglia+2008}. We
detect a moderately significant gradient of \(4.3\pm1.3\,\kmsdeg\) at a
position angle of \(-149_{-13}^{+17}\) degrees, similar to past
detections
\citetext{\citealp[e.g.,][]{arroyo-polonio+2024}; \citealp{battaglia+2008}; \citealp[but
see also][]{strigari2010}; \citealp{martinez-garcia+2023}}.

Fig.~\ref{fig:scl_velocity_gradient_scatter} plots the combined velocity
sample and the MCMC samples for the velocity gradient. While some
samples have a gradient consistent with 0, most samples have a positive
velocity gradient, consistent with the rolling median trend. The
velocity gradient in Scl is misaligned with the proper motion.

@@

\emph{Binarity}. Binary star systems may inflate the inferred system's
velocity dispersion. For Scl and UMi with high measured velocity
dispersion, binaries likely add \(\sim 1\,\kms\) to the velocity
dispersion \citep{spencer+2018, gration+2025}.

\emph{Multiple populations}. Both Sculptor and Ursa Minor likely contain
multiple populations with different kinematics (see
Section~\ref{sec:peculiarities}). We do not model these separately---it
is unclear how to define a velocity dispersion in such a case.

@@

\emph{Tangential motions.} Precise star-by-star proper motions would
further test for kinematic disequilibrium. Presently, \emph{Gaia}
uncertainties are too large to permit such studies for Sculptor and Ursa
Minor.\footnote{Specifically, typical proper motion uncertainties for
  faint, member stars of Scl and UMi are around \(0.5\,\masyr\),
  corresponding to velocities \(\sim 100\,\kms\).}

@@

In this section, we analyze literature samples of LOS velocity
measurements for both Scl and UMi. In each case, we find systemic
velocities and dispersions consistent with past work. We detect weak
evidence for a velocity gradient in Scl, but the gradient is misaligned
with the proper motion of Scl, so it is unlikely to be of tidal origin.
Sculptor also shows evidence for a mildly rising velocity dispersion
with radius; however, the outer regions appear to be flat in velocity
dispersion. The lack of observational evidence for ongoing tidal
disruption in the velocity distribution of stars in Scl and UMi further
supports our interpretations in the main text.
